Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-18 probabilisticProgrammingMartingales.tex
Sec. 3: ongoing

We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_n = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
It follows that $g \geq f$ as required.

@@

\subsection{Partial ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it. Take, for example, the term
\begin{align*}
&(\tY \lambda f, n: \\
&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
&) \, \underline{0}
\end{align*}
which generates a geometric distribution.
    Despite its simplicity, its $Rch$ contains all the terms
\begin{enumerate}
    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\underline{i}$
    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
\end{enumerate}
\akr{TODO: make this into a diagram with arrows.}

@@

\lo{So we have:
\[
\begin{array}{ll}
1 \to 11 \to 4 \to 5 \to 6 \to 7 \to 8\\
7 \to 9 \to 11 \to \cdots
\end{array}
\]
You have 2, 3 and 10 because your $\tY$-redex rule is
\[
\tY \lambda x. s \to \lambda z. (\lambda x.s)\, (\tY \lambda x. s) \, z
\]
which is not the official definition.}

@@

\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n. n \oplus_{0.5} f(n+1)) \underline 0$. It is then easy to see that the following is a partial ranking function:
\begin{align*}
(\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2 \\
\underline i \oplus_{0.5} (\tY f, n. n \oplus_{0.5} f(n+1)) (\underline i + 1) \mapsto 1 \\
\underline i \mapsto 0
\end{align*}

>>>>

We have $\mathbb E[f(M_n)] = \sum_{k=n}^\infty \mathbb P[M_k = E[\tY N] \text{ for some }E, N] \to 0 \text{ as } n \to \infty$; 
therefore as $g(M) - f(M) < 0$, eventually $\mathbb E[g(M_n)] < 0$, which is impossible. 
It follows that $g \geq f$ as required.

@@

\subsection{Partial ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.

\begin{example}[Geometric distribution]
Consider the term
\begin{align*}
&(\tY \lambda f, n: \\
&\quad \tif{\tsample - 0.5 < 0}{n}{f (n+1)} \\
&) \, \underline{0}
\end{align*}
which generates a geometric distribution.
    Despite its simplicity, its $Rch$ contains all the terms
\begin{enumerate}
    \item $(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{0}$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) \underline{i}$
    \item $(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) \underline{i}$
    \item $(\lambda n: \tif{\tsample - 0.5 < 0}{n}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{n}{f (m+1)}) (m+1)}) \underline{i}$
    \item $\tif{\tsample - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline r - 0.5 < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\tif{\underline{r - 0.5} < 0}{\underline{i}}{(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)}$
    \item $\underline{i}$
    \item $(\tY \lambda f, m: \tif{\tsample - 0.5 < 0}{m}{f (m+1)}) (\underline{i}+1)$
    \item $(\lambda z.(\lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) (\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)}) z) (\underline{i} + 1)$.
\end{enumerate}
\akr{TODO: make this into a diagram with arrows.}

@@

\lo{So we have:
\[
\begin{array}{ll}
1 \to 11 \to 4 \to 5 \to 6 \to 7 \to 8\\
7 \to 9 \to 11 \to \cdots
\end{array}
\]
You have 2, 3 and 10 because your $\tY$-redex rule is
\[
\tY \lambda x. s \to \lambda z. (\lambda x.s)\, (\tY \lambda x. s) \, z
\]
which is not the official definition.}
\end{example}

@@

\paragraph{}
With this abbreviation, the geometric distribution example from earlier can be written as $(\tY f, n. n \oplus_{0.5} f(n+1)) \, \underline 0$. It is then easy to see that the following is a partial ranking function:
\begin{align*}
(\tY f, n. n \oplus_{0.5} f(n+1)) N \mapsto 2 \\
\underline i \oplus_{0.5} (\tY f, n. n \oplus_{0.5} f(n+1)) (\underline i + 1) \mapsto 1 \\
\underline i \mapsto 0
\end{align*}

### 2020-12-18 ranking-supermartingales.tex
Sec. 3: ongoing

\begin{definition}\rm
A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $\expect{|Y_n|} < \infty$, and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$.
\lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
%For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where 
\(T(s) := \min \set{n \mid Y_n(s) = 0}\).
It is straightforward to see that $T$ is a stopping time w.r.t.~$(\calF_n)_{n \in \omega}$.}
A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
\end{definition}

@@

\begin{theorem}[Soundness]
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.

>>>>

\begin{definition}\rm
A sequence of random variables $(Y_n)_{n \in \omega}$ adapted to a filtration $(\calF_n)_{n \in \omega}$ is a \emph{supermartingale} if for all $n \in \omega$, $\expect{|Y_n|} < \infty$, and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n$.
\lo{The preceding means: $\forall A \in \calF_n$, $\int_A \dif \mu \, Y_{n+1} \leq \int_A \dif \mu \,Y_n$.}
%For a stopping time $T$ and $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$.
\changed[lo]{For $\epsilon > 0$, it is a \emph{$\epsilon$-ranking supermartingale} if, in addition, for all $n$, $Y_n \geq 0$ and $\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon \cdot \mathbf{1}_{\set{T > n}}$ where the random variable $T$ is defined as
\(T : s \mapsto \min \set{n \mid Y_n(s) = 0}\).
(It is straightforward to see that $T$ is a stopping time w.r.t.~$(\calF_n)_{n \in \omega}$.)}
A \emph{ranking supermartingale} is a $\epsilon$-ranking supermartingale for some $\epsilon > 0$.
\end{definition}

@@

\begin{theorem}[Soundness] \label{thm:rankable implies termination}
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $T^{\tY}_M < \infty$ a.s.~(equivalently $M$ is AST), and $\expect{T^{\tY}_M} < \infty$.
